---
title: "Diabetes: High Variance and Descriptive Statistics"
author: "Bianca Greul"
date: "15 5 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r libraries, include=FALSE}
# Bioconductor
library(affy)
library(vsn)
library(AnnotationDbi)
library(hgu133plus2hsenstcdf)
library(hgu133plus2hsenstprobe)
library(hexbin)

library(rstudioapi)
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(stats)
```

Diabetes type 1 dataset, GSE53454, data is downloaded from the GEO database

```{r}

# load necessary files from our created RData.

projectPath <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(paste(projectPath, "sessions", "RData", sep = "/"))

load(file = "diabetes_vsnrma_matrix.RData")

load(file = "diabetes_GeneExprs.RData")

load(file = "diabetes_GeneExprs_sub.RData")

load(file = "diabetes_GeneExprs_sub_combined.RData")

load(file = "diabetes_GeneExprs_combined.RData")

load(file = "diabetes_TRAexprs.RData")
```


# 1. Microarray Analysis

## 1.1 Chips for further analysis

```{r}

diabetes_ctrl <- diabetes_vsnrma_matrix[, 1:13]
diabetes_treat <- diabetes_vsnrma_matrix[, 14:24]

## The Control group has two additional chips. That might lead to problems down the line. Therefore, we might have to removed those two chips.

col_diab <- c(2:11, 13)
diabetes_ctrl_same_chips <- diabetes_vsnrma_matrix[, col_diab]

## We discussed removing two chips due to unsatisfactory results in the quality control. Here are control and treatment group with the two chips and those linked to each other removed.
col_ctrl <- c(3, 5:8, 10:13)
diabetes_crtl_finished <- diabetes_vsnrma_matrix[, col_ctrl]

col_treat <- c(14, 16, 18:24)
diabetes_treat_finished <- diabetes_vsnrma_matrix[, col_treat]

```


## 1.2 High variance genes

### Top 5% of variance

```{r}

## Calculate variance of rows (genes)

var_diab <- apply(diabetes_GeneExprs, 1, var, na.rm=TRUE)

## Create function to filter out the top 5% of variance.

top5 <- function(x) {
  x >= quantile(x, 0.95)
}

## Apply function to variance set, to determine the top 5% of variance.

top5_diab <- top5(var_diab)

top5_diab_names <- which(top5_diab == "TRUE")

high_var_diabetes <- diabetes_GeneExprs[top5_diab_names, , drop = FALSE]

dim(high_var_diabetes)

## data set complete: 94451 genes
## data set high variance: 4729 genes
```


## 1.3 Median gene expression

### 1.3.1 Median expression of genes that is more than twice that of the general median.

```{r}
## Search for genes, whose median gene expression are more than 1.9 times that of the median.

med_diab <- apply(diabetes_GeneExprs, 1, median, na.rm=TRUE)

## Names of genes that fit this criteria: 8 genes
## GCG, INS-IGF2, INS, PRSS2, MT-ND3, RPL31, PPL37A, SST

which(med_diab > (1.9 * median(med_diab)))

## Expression values of those genes.

diab_above2med <- diabetes_GeneExprs[which(med_diab > (1.9 * median(med_diab)))]
diab_above2med
```

### 1.3.2 Median gene expression comparison treatment vs control

```{r}
## separate Control and Treatment group to compare median gene expression levels.

median_ctrl <- apply(diabetes_ctrl, 1, median, na.rm=TRUE)
median_treat <- apply(diabetes_treat, 1, median, na.rm=TRUE)


## Search for genes, whose median gene expression have been increased 1.5 times during treatment in comparison to control

median_1_5 <- which(median_treat > (1.5 * median_ctrl))
median_1_5

## This is true for the following genes: CXCL9 CXCL10 CXCL11 IDO1

## Their gene expression changed from...
median_ctrl[median_1_5]
## ... to:
median_treat[median_1_5]

## Search for genes in treatment group, whose median gene expression is reduced by 1.2 times compared to the control group

median_reduced <- which((1.2 * median_treat) < (median_ctrl))
head(median_reduced)

## Following genes have a reduced median gene expression after treatment: OLFM4, AKR1B10

```


## 1.4 Variance

### Variance of Microarrays: Control vs Treatment

```{r}
## Expected results: Control Microarrays only have a small variance compared to Treatment Microarrays with a higher variance.

var_ctrl <- apply(diabetes_ctrl, 1, var, na.rm=TRUE)
var_treat <- apply(diabetes_treat, 1, var, na.rm=TRUE)

## Control should have a smaller variance than treatment group. Length should be small.
var_t_smaller_c <- which(var_treat < var_ctrl)
length(var_t_smaller_c)

## Treatment should have a larger variance than control group. Length should be large.
var_t_greater_c <- which(var_treat > var_ctrl)
length(var_t_greater_c)

## Surprising results: Only roughly five ninth of the genes have a higher variance of their gene expression after treatment.
```


## 1.5 Checking normality

```{r}
qqnorm(diabetes_ctrl, main = "Normal QQ-Plot of Diabetes Control", ylab = "Diabetes Control"); qqline(diabetes_ctrl)

qqnorm(diabetes_treat, main = "Normal QQ-Plot of Diabetes Treatment", ylab = "Diabetes Treatment"); qqline(diabetes_treat)

## Control and treatment group are not normally distributed.

```


# 2. TRA Analysis 

Steps already taken in Analysis:

```{r}
setwd(paste(projectPath, "rawdata", "tables" ,sep = "/"))
pancreas_specific_genes = read.csv("pancreas_specific_genes_human.csv")
pancreas_specific_genes = sort(pancreas_specific_genes)


# create a expression matrix with only those genes
diabetes_GeneExprs_sub = diabetes_GeneExprs[which(rownames(diabetes_GeneExprs) %in% pancreas_specific_genes),]

dim(diabetes_GeneExprs_sub) 
# 1569   24 
length(unique(rownames(diabetes_GeneExprs_sub)))
# 250 (number of TRA genes in the expression matrix)
head(diabetes_GeneExprs_sub)

## create a expression matrix with all TRA genes

setwd(paste(projectPath, "rawdata", "tables" ,sep = "/"))
all_TRA_genes = read.csv("TRA_genes_human.csv")
all_TRA_genes = sort(all_TRA_genes)

all_TRA_GeneExprs_sub = diabetes_GeneExprs[which(rownames(diabetes_GeneExprs) %in% all_TRA_genes),]

dim(all_TRA_GeneExprs_sub) 
#
length(unique(rownames(all_TRA_GeneExprs_sub)))
#  (number of TRA genes in the expression matrix)
head(all_TRA_GeneExprs_sub)
```


## 2.1 pancreatic TRA expression analyis of all diabetis sets

```{r}
diabetes_GeneExprs_sub_combined <- combineGeneExprs_sum(diabetes_GeneExprs_sub)

head(diabetes_GeneExprs_sub_combined)
dim(diabetes_GeneExprs_sub_combined)
##250
```

```{r}
### mean insgesamt und für jedes Set einzelnd
mean(diabetes_GeneExprs_sub_combined)
## 8.79089
apply(diabetes_GeneExprs_sub_combined, 2, mean)

### median insgesamt und für jedes Set einzelnd
median(diabetes_GeneExprs_sub_combined)
## 7.996373
apply(diabetes_GeneExprs_sub_combined, 2, median)

### sd ingesamt und für jedes Set einzelnd
sd(diabetes_GeneExprs_sub_combined)
##  1.6099
apply(diabetes_GeneExprs_sub_combined, 2, sd)

```

```{r}
## lowest expressed Gene each chip:

for(i in 1:24)
{min_value <- min(diabetes_GeneExprs_sub_combined[,i])
min_gene <- row.names(diabetes_GeneExprs_sub_combined)[diabetes_GeneExprs_sub_combined[,i] == min_value]
print(min_value)
print(min_gene)}
```

```{r}
### highest expressed Gene each chip:

for(i in 1:24)
{max_value <- max(diabetes_GeneExprs_sub_combined[,i])
max_gene <- row.names(diabetes_GeneExprs_sub_combined)[diabetes_GeneExprs_sub_combined[,i] == max_value]
print(max_value)
print(max_gene)}

```

## 2.1.5 all TRA expression analyis of all diabetes sets

```{r}
all_TRA_GeneExprs_sub_combined <- combineGeneExprs_sum(all_TRA_GeneExprs_sub)

head(all_TRA_GeneExprs_sub_combined)
dim(all_TRA_GeneExprs_sub_combined)
##
```
```{r}
### mean insgesamt und für jedes Set einzelnd
mean(all_TRA_GeneExprs_sub_combined)
## 8.79089
apply(all_TRA_GeneExprs_sub_combined, 2, mean)

### median insgesamt und für jedes Set einzelnd
median(all_TRA_GeneExprs_sub_combined)
## 7.996373
apply(all_TRA_GeneExprs_sub_combined, 2, median)

### sd ingesamt und für jedes Set einzelnd
sd(all_TRA_GeneExprs_sub_combined)
##  1.6099
apply(all_TRA_GeneExprs_sub_combined, 2, sd)

```

```{r}
## lowest expressed Gene each chip:

for(i in 1:24)
{min_value <- min(all_TRA_GeneExprs_sub_combined[,i])
min_gene <- row.names(all_TRA_GeneExprs_sub_combined)[all_TRA_GeneExprs_sub_combined[,i] == min_value]
print(min_value)
print(min_gene)}
```
```{r}
### highest expressed Gene each chip:

for(i in 1:24)
{max_value <- max(all_TRA_GeneExprs_sub_combined[,i])
max_gene <- row.names(all_TRA_GeneExprs_sub_combined)[all_TRA_GeneExprs_sub_combined[,i] == max_value]
print(max_value)
print(max_gene)}
```


## 2.2 High Variance TRA genes in diabetes data set

```{r}
## Calculate variance of rows (genes)

var_di <- apply(diabetes_GeneExprs_sub_combined, 1, var, na.rm=TRUE)

## Create function to filter out the top 25% of variance.

top5 <- function(x) {
  x >= quantile(x, 0.75)
}

## Apply function to variance set, to determine the top 5% of variance.

top5_di <- top5(var_di)

top5_di_names <- which(top5_di == "TRUE")

high_var_diabetes_sub <- diabetes_GeneExprs_sub_combined[top5_di_names, , drop = FALSE]

dim(high_var_diabetes_sub)

## data set complete: 250 genes
## data set high variance: 63 genes
```

#### Boxplot of 25% high variance genes

```{r}
par(las = 2)
par(mai = c(0.7,0.4,0.5,0.1))

boxplot(t(high_var_diabetes_sub), col = rainbow(63), cex.axis=0.5, main = "Top 25% high variance genes of diabetes data set"); 
abline(h = median(diabetes_GeneExprs_sub_combined), col = "red")
```


## 2.3 TRAs in Diabetes Control Group 

### Read in functions for combining gene expression: median or sum.

```{r}
## Create new matrix containing the control groups of the diabetes dataset with only TRA specific genes. We are unsure whether we should choose sum or median to combine the gene expressions.

set.seed(977)

## Median

diabetes_TRA_ctrl_median <- diabetes_GeneExprs_sub_combined[, 1:13]

## Sum

diabetes_GeneExprs_sub_combined_sum <- combineGeneExprs_sum(diabetes_GeneExprs_sub)
diabetes_TRA_ctrl_sum <- diabetes_GeneExprs_sub_combined_sum[, 1:13]
```


### 2.3.1 Genes with high expression levels 

#### Results are not similar for sum/median technique

```{r}
# Median

med_diab_TRA_median <- apply(diabetes_TRA_ctrl_median, 1, median, na.rm=TRUE)
median(med_diab_TRA_median)

## Median: 8.068697

## Search for genes, whose median gene expression are more than 1.8 times that of the median.
which(med_diab_TRA_median > (1.8 * median(med_diab_TRA_median)))
## Names of genes that fit this criteria: 2 in total.
## INS, PRSS2

## Expression values of those genes:
diab_above2med_TRA_median <- diabetes_GeneExprs[which(med_diab_TRA_median > (1.8 * median(med_diab_TRA_median)))]
diab_above2med_TRA_median
```

```{r}
# Sum 

med_diab_TRA_sum <- apply(diabetes_TRA_ctrl_sum, 1, median, na.rm=TRUE)
median(med_diab_TRA_sum)

## Median: 38.43985

## Search for genes, whose median gene expression are more than 1.5 times that of the median.
which(med_diab_TRA_sum > (5 * median(med_diab_TRA_sum)))
## Names of genes that fit this criteria: 2 in total (some are doubled).
## ABCC8, CD44, FGFR1, GNAS, IL32, P4HB, PABPC4, PSEN2, RPS9, SERPINB6, TPM1, VEGFA

## Expression values of those genes:
diab_above2med_TRA_sum <- diabetes_GeneExprs[which(med_diab_TRA_sum > (5 * median(med_diab_TRA_sum)))]
diab_above2med_TRA_sum
```


### 2.3.2 Genes that are differently expressed over the time period

#### Results are the same for sum/median technique

```{r}
# Median 

## Genes that are expressed 1.3 times less in the last chip than in the first.
diabetes_TRA_ctrl_median_receding <- (which(diabetes_TRA_ctrl_median[,1] > (1.3 * diabetes_TRA_ctrl_median[,13])))
## 10 genes

## Genes that are expressed 1.2 times more in the last chip than in the first.
diabetes_TRA_ctrl_median_growing <- (which((1.2 *diabetes_TRA_ctrl_median[,1]) < diabetes_TRA_ctrl_median[,13]))
## 4 genes

diabetes_TRA_ctrl_median_receding
diabetes_TRA_ctrl_median_growing

## Is the change significant?
## Only 10 and 4 genes differently expressed at threshold levels of 1.3 and 1.2
```

```{r}
# Sum

## Genes that are expressed 1.3 times less in the last chip than in the first.
diabetes_TRA_ctrl_sum_receding <- (which(diabetes_TRA_ctrl_sum[,1] > (1.3 * diabetes_TRA_ctrl_sum[,13])))
## 10 genes

## Genes that are expressed 1.2 times less in the last chip than in the first.
diabetes_TRA_ctrl_sum_growing <- (which((1.2 *diabetes_TRA_ctrl_sum[,1]) < diabetes_TRA_ctrl_sum[,13]))
## 4 genes

diabetes_TRA_ctrl_sum_receding
diabetes_TRA_ctrl_sum_growing

## Is the change significant?
## Only 10 and 4 genes differently expressed at threshold levels of 1.3 and 1.2
```


### 2.3.3 What are the genes that have a high variance: Important or house keeping genes?

```{r}
## Calculate variance of rows (genes)

var_diabetes_TRA_ctrl_median <- apply(diabetes_TRA_ctrl_median, 1, var, na.rm=TRUE)

## Create function to filter out the top 2.5% of variance.

top2.5 <- function(x) {
  x >= quantile(x, 0.975)
}

## Apply function to variance set, to determine the top 5% of variance.

top2.5_var_ctrl_median <- top2.5(var_diabetes_TRA_ctrl_median)

top2.5_var_ctrl_median_names <- which(top2.5_var_ctrl_median == "TRUE")

high_var_diabetes_ctrl_median <- diabetes_TRA_ctrl_median[top2.5_var_ctrl_median_names, , drop = FALSE]

dim(high_var_diabetes_ctrl_median)

## data set complete: 250 genes
## data set high variance: 7 genes

rownames(high_var_diabetes_ctrl_median)

# High variance genes and their function:

## CELA2B: serine proteases that hydrolyze elastin; secreted from pancreas as zymogen
## CELP: carboxyl ester lipase pseudogene
## CUZD1: CUZD1 antiserum inhibits cell attachment and proliferation of ovarian cancer cells so may be involved in these processes.
## GP2: encodes an integral membrane protein; antigen binding
## PNLIP: encodes a member of the lipase family of proteins
## PNLIPRP1: annotations related to this gene include calcium ion binding and triglyceride lipase activity

# Might be important

## SPP1: involved in the attachment of osteoclasts to the mineralized bone matrix; cytokine that upregulates expression of interferon-gamma and interleukin-12

```

#### Boxplot of identified 2.5% high variance genes in diabetes control

```{r}
par(las = 2)
par(mai = c(1.1,0.5,0.5,0.1))

boxplot(t(high_var_diabetes_ctrl_median), col = rainbow(7), main = "2.5% high variance genes of diabetes control group"); 
abline(h = median(med_diab_TRA_median), col = "red")
```


## 2.4 Comparison of control and treatment group

### 2.4.1 Find high variance genes of treatment group

```{r}
## Calculate variance of rows (genes) of the diabetes treatment group

diabetes_TRA_treat_median <- diabetes_GeneExprs_sub_combined[, 14:24]
var_diabetes_TRA_treat_median <- apply(diabetes_TRA_treat_median, 1, var, na.rm=TRUE)

## Create function to filter out the top 2.5% of variance.

top2.5 <- function(x) {
  x >= quantile(x, 0.975)
}

## Apply function to variance set, to determine the top 5% of variance.

top2.5_var_treat_median <- top2.5(var_diabetes_TRA_treat_median)

top2.5_var_treat_median_names <- which(top2.5_var_treat_median == "TRUE")

high_var_diabetes_treat_median <- diabetes_TRA_treat_median[top2.5_var_treat_median_names, , drop = FALSE]

dim(high_var_diabetes_treat_median)

## data set complete: 250 genes
## data set high variance: 7 genes

rownames(high_var_diabetes_treat_median)

```

```{r}
med_diab_TRA_treat <- apply(diabetes_TRA_treat_median, 1, median, na.rm=TRUE)

par(las = 2)
par(mai = c(1.1,0.5,0.5,0.1))

boxplot(t(high_var_diabetes_treat_median), col = rainbow(7), main = "2.5% high variance genes of diabetes treatment group"); 
abline(h = median(med_diab_TRA_median), col = "red")
```

### 2.4.2 Are there genes that are present in both 2.5% high variance control and 2.5% high variance treatment?

```{r}
rownames(high_var_diabetes_ctrl_median)
rownames(high_var_diabetes_treat_median)

## Same genes: CUZD1, GP2, PNLIPRP1, PNLIP

## Create a data set with those genes: ctrl and treatment.
## Control data set reduced to the same number of microarrays as treatment.

col_diab <- c(2:11, 13)

ctrl_treat_high_var <- cbind(high_var_diabetes_ctrl_median["CUZD1",col_diab], high_var_diabetes_treat_median["CUZD1",],
                            high_var_diabetes_ctrl_median["GP2",col_diab], high_var_diabetes_treat_median["GP2",],
                            high_var_diabetes_ctrl_median["PNLIPRP1",col_diab], high_var_diabetes_treat_median["PNLIPRP1",], 
                            high_var_diabetes_ctrl_median["PNLIP",col_diab], high_var_diabetes_treat_median["PNLIP",]);
colnames(ctrl_treat_high_var) <- c("CUZD1 ctrl","CUZD1 treat", "GP2 ctrl", "GP2 treat", "PNLIPRP1 ctrl", "PNLIPRP1 treat", "PNLIP ctrl", "PNLIP treat")

## Boxplot

par(las = 2)
par(mai = c(1.1,0.5,0.5,0.3))

boxplot(ctrl_treat_high_var, col = rainbow(8), main = "High variance genes present in the top 2.5% in both control and treatment", cex.axis = 0.7);
abline(h = median(diabetes_GeneExprs_sub_combined), col = "red")

```

### 2.4.3 High variance gene in both control and treatment group comparison

```{r}

high_var_diabetes_ctrl_ordered <- high_var_diabetes_ctrl_median[, c(1, 3, 5, 8, 12, 2, 4, 6, 7, 9, 10, 11, 13)]

ctrl_high_var <- cbind(high_var_diabetes_ctrl_ordered["CUZD1",], high_var_diabetes_ctrl_ordered["GP2",],
                       high_var_diabetes_ctrl_ordered["PNLIPRP1",], 
                       high_var_diabetes_ctrl_ordered["PNLIP",])

high_var_diabetes_treat_ordered <- high_var_diabetes_treat_median[, c(3, 5, 1, 2, 4, 6, 7, 8, 9, 10, 11)]

treat_high_var <- cbind(high_var_diabetes_treat_ordered["CUZD1",], high_var_diabetes_treat_ordered["GP2",],
                        high_var_diabetes_treat_ordered["PNLIPRP1",], high_var_diabetes_treat_ordered["PNLIP",])

col_ctrl_treat <- c("CUZD1", "GP2", "PNLIPRP1", "PNLIP")

# Plots

par(mfrow = c(2,2), mai = c(0.75,0.7,0.5,0.1))

## control 

for (i in 1:ncol(ctrl_high_var)){
  plot(ctrl_high_var[,i], 
       type = "b", 
       xlab = "Microarray Number", ylab = "Gene expression", 
       main = paste(col_ctrl_treat[i], "\n", "Control"));
  axis(1,  at = seq(1, nrow(ctrl_high_var), by=1))
}

## treatment

for (i in 1:ncol(treat_high_var)){
  plot(treat_high_var[,i], 
       type = "b", 
       xlab = "Microarray Number", ylab = "Gene expression", 
       main = paste(col_ctrl_treat[i], "\n", "Treatment"));
}
axis(1, at = seq(1, nrow(treat_high_var), by=1))

```


# 3. Pancreas

```{r}
# load necessary files from our created RData.

projectPath <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(paste(projectPath, "sessions", "RData", sep = "/"))

load(file = "pancreas_vsnrma_matrix.RData")

load(file = "pancreas_GeneExprs.RData")

load(file = "pancreas_GeneExprs_sub.RData")

load(file = "pancreas_GeneExprs_sub_combined.RData")

load(file = "pancreas_GeneExprs_combined.RData")

load(file = "pancreas_TRAexprs.RData")
```

```{r}
## Calculate variance of rows (genes) of the pancreas control group

pancreas_TRA_ctrl <- pancreas_GeneExprs_sub_combined[, 4:6]
var_pancreas_TRA_ctrl <- apply(pancreas_TRA_ctrl, 1, var, na.rm=TRUE)

## Create function to filter out the top 2.5% of variance.

top2.5 <- function(x) {
  x >= quantile(x, 0.975)
}

## Apply function to variance set, to determine the top 5% of variance.

top2.5_var_ctrl_pancreas <- top2.5(var_pancreas_TRA_ctrl)

top2.5_var_ctrl_pancreas_names <- which(top2.5_var_ctrl_pancreas == "TRUE")

high_var_ctrl_pancreas <- pancreas_TRA_ctrl[top2.5_var_ctrl_pancreas_names, , drop = FALSE]

dim(high_var_ctrl_pancreas)

## data set complete: 250 genes
## data set high variance: 7 genes

rownames(high_var_ctrl_pancreas)

```

#### Boxplot of identified 2.5% high variance genes in pancreas control

```{r}

med_pancreas_TRA <- apply(pancreas_TRA_ctrl, 1, median, na.rm=TRUE)

par(las = 2)
par(mai = c(1.1,0.5,0.5,0.1))

boxplot(t(high_var_ctrl_pancreas), col = rainbow(7), main = "2.5% high variance genes of pancreas control group"); 
abline(h = median(med_pancreas_TRA), col = "red")
```

#### High variance genes depending on the chip

```{r}
ctrl_high_var_pancreas <- cbind(high_var_ctrl_pancreas["FBXL8",], 
                                high_var_ctrl_pancreas["GOLGA8A",], 
                                high_var_ctrl_pancreas["GOLGA8B",], 
                                high_var_ctrl_pancreas["IL32",], 
                                high_var_ctrl_pancreas["PRSS2",], 
                                high_var_ctrl_pancreas["SEL1L",], 
                                high_var_ctrl_pancreas["SYCN",])

col_ctrl_treat_pancreas <- c("FBXL8", "GOLGA8A", "GOLGA8B", "IL32", "PRSS2", "SEL1L", "SYCN" )

# Plots

par(mfrow = c(2,2), mai = c(0.75,0.7,0.5,0.1))

## control 

for (i in 1:ncol(ctrl_high_var_pancreas)){
  plot(ctrl_high_var_pancreas[,i], 
       type = "b", 
       xlab = "Microarray Number", ylab = "Gene expression", 
       main = paste(col_ctrl_treat_pancreas[i], "\n", "Pancreas Control"))
}
```

